Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-05 paper.qmd
brain-melting nodes and equator plane intersections

<!-- as a minimum, we should say something about the WHAT -->
In general, the method for estimating the risk to commercial aircraft
from reentering debris objects revolves around estimating the exposure
of a single flight and the population of flights to the re-entering
object. In that respect, a robust determination of flight density is a
key aspect.

We follow the same mathematical approach of [@patera2008;
@wrightAirspaceClosuresDue2025]. However, this paper augments the
approximation of risk exposure. Instead of approximating flight density
with population density [@patera2008] or aircraft transponder data
[@wrightAirspaceClosuresDue2025], we use actual flown flight
trajectories.

@@

To build an upper threshold for flight exposure, this study uses the
busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.
<!-- does ISO do here anything? --> The flight densities are
characterised on an hourly basis for all days in these weeks.

@@

<!-- next sentence feels like an orphan ... do we need this here? ------>
Next sections will develop the mathematical background for the
evaluation of collision and casualties risk of which traffic density is
a key component.

@fig-density-summer2024-res-3 shows hourly flight density for the 5th of
July 2024. The chart shows the expected outcomes in terms of major
traffic flows, and areas of high or low traffic. For example, cells with
a density of zero are clearly visible for areas of war or no-fly zones
(Ukraine, Libya or Syria). The major traffic flows (North Atlantic,
Canaries and East-bound over Türkiye) are clearly highlighted showing
densities of more than 10.

@@

Our approach builds on the discretisation of the geospatial risk area.
For this purpose we use a hexagon-based tessellation pattern.
Uber's H3 method ensures non-overlapping hexagons (and 5 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
Using H3 to discretise the area of interest $a$ gives

@@

Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations:
<!-- TERMINOLOGY? rectangular solid -- impact top plane, front plane? Does the literature use "view"  -->

@@

Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes.
The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
At the same time, the falling of the re-entering debris object at regular en-route flight levels will be vertical[^ballistic_descent].
This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as

@@

These aircraft properties are well-known. 
While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
The values $W_t$, $H_t$, $L_t$ and $C_{t}$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used $\varepsilon_t = \qty{1000}{m^2}$ missing entries.

@@

Debris that re-enter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to re-entry even if initially on elliptical
ones.[^circular-orbit]

The angular position, $\theta = \omega + \nu$, of the re-entry point from the
ascending node, see @sec-orbitals, is uniformly distributed for uncontrolled
re-entries, $\sigma(\theta) = 1 / (2\pi)$.
This is due to uncertainties in atmospheric drag (among other things solar
activity can expand the atmosphere) and the gravitational effects of the
Equatorial bulge.
Hence we assume that the impact longitude, $\lambda$, is uniformly distributed.

>>>>

<!-- as a minimum, we should say something about the WHAT -->
In general, the method for estimating the risk to commercial aircraft
from re-entering debris objects revolves around estimating the exposure
of a single flight and the population of flights to the re-entering
object. In that respect, a robust determination of the flight density for encounter airspace volumes is a
key aspect.

We follow the same mathematical approach of [@patera2008;
@wrightAirspaceClosuresDue2025]. However, this paper augments the
approximation of risk exposure. Instead of approximating flight density
with population density [@patera2008] or aircraft transponder data
[@wrightAirspaceClosuresDue2025], we use actual flown flight
trajectories. This provides a more realistic figure of the number of affected air traffic.
Thus, flight density is directly related to the number of flight operations per volume of airspace.

@@

To build an upper threshold for the flight exposure, this study uses the
busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.
This ensures a reasonable value, as peak airspace utilisation provides a handle on the air traffic operations and interactions of the different aircraft as part of wider *'traffic flows'*.
<!-- does ISO do here anything? --> The flight densities are
characterised on an hourly basis for all days in these weeks per H3-cell.

@@

<!-- next sentence feels like an orphan ... do we need this here? ------>
The next sections will develop the mathematical background for the
evaluation of the collision and casualty risk for which air traffic density is
a key component.

@fig-density-summer2024-res-3 shows the hourly flight density for the 5th of
July 2024. The chart depicts the expected outcomes in terms of major
traffic flows, and areas of high or low traffic. For example, cells with
a density of zero are clearly visible for areas of war or no-fly zones
(Ukraine, Libya or Syria). At the same time, the major traffic flows (North Atlantic,
Canaries and East-bound over Türkiye) are clearly highlighted showing
densities of more than 10.

@@

Our approach builds on the discretisation of the geospatial risk area.
For this purpose we use a hexagon-based tessellation pattern.
Uber's H3 method ensures non-overlapping hexagons (and 5 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
While this paper focusses on the determination of the risk exposure per cell, the approach would allow to identify adjacent cells to account for the inherent uncertainty and risk appetite of decision-makers.

Using H3 to discretise the area of interest $a$ gives

@@

Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations:

@@

Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes (i.e., flight levels).
The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
At the same time, the trajectory of a falling re-entering debris object at regular en-route flight levels will be a vertical decent movement[^ballistic_descent].
This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as

@@

These aircraft properties are well-known design parameters. 
While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
The values $W_t$, $H_t$, $L_t$ and $C_{t}$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used $\varepsilon_t = \qty{1000}{m^2}$ missing entries or unknown aircraft types.

@@

Debris that re-enter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to re-entry even if initially on elliptical
ones.[^circular-orbit]
A characteristic point for the determination of the re-entry position is the *'ascending node'*.
This denotes the intersection point of the object crossing the plane of Earth's equator if moving from the southern hemnisphere to the nothern one.
The respective angular position, $\theta = \omega + \nu$, of the re-entry point from the
ascending node, see @sec-orbitals, is uniformly distributed for uncontrolled
re-entries, $\sigma(\theta) = 1 / (2\pi)$.
This is due to uncertainties in atmospheric drag (among other things solar
activity can expand the atmosphere) and the gravitational effects of the
Equatorial bulge.
Hence we assume that the impact longitude, $\lambda$, is uniformly distributed.
